---
title: "206 tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r test}
dream_team <- Cheyenne + Katie + Emily
```

```{r library packages}
library(ape)
library(phytools)
library(geiger)
```


```{tree visualizations}
#http://www.phytools.org/Cordoba2017/ex/15/Plotting-methods.html
```
```{r tree analysis}
#analysis of similarity using continuous variables
#https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.2051
```
```{building a tree}
#https://cran.r-project.org/web/packages/data.tree/vignettes/data.tree.html
```



TRYING THINGS
```{r read data}
data <- read.csv("trait_data.csv", header= TRUE, sep= ",")
```


```{r install ggtree}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ggtree", version = "3.8")
library(ggtree)
```

```{r read table tree}
library(data.tree)
library(data.table)
data_2 <- as.data.frame(data)
data_3 <- ?as.Node.data.frame(data_2)
tree <- read.tree(data_2)
```
https://cran.r-project.org/web/packages/data.tree/vignettes/data.tree.html


Making a tree based on the character table 
```{r tree text}
text_string <- "((((lamprey, sea bass),antelope), (bald eagle), alligator));"
```


```{r reading tree}
treew<- read.tree(text=text_string)
```

```{r plotting tree}
plot(treew,no.margin=TRUE, edge.width=2)
```

geiger stuff
pg 29 of vignette
```{r geiger data}
data(caudata)
data("whales")

View(caudata$dat)
```

Maybe check out phytools:
https://cran.r-project.org/web/packages/phytools/phytools.pdf

Hey guys, below is just a start on what I think we should include if we do proceed with using ggtree and a newick file. Hopefully that will be the direction we go in!

I think a good project outline would be:
brief review of navigating in r (we should look at what 205 is doing in their tutorial)
importing newick files and setting a working directory
ggplot and ggtree
making different visuals from the same tree data
interpreting a tree basics
customizing a tree to show different information

```{r}
#install neccessary packages..may take a few minutes because installing ggtree will also install all neccessary supplementary files (there is a masking function involved with installing ggtree that will hide the installation of supplementary packages)
install.packages("tidyverse")
library(tidyverse)
library(ggtree)
#prior to do the below commands I downloaded a newick file and placed it in the working directory
tree <- read.tree("tree_newick(1).nwk") #imports the newick file into R
tree #gives a nice summary of the newick file and data stored in the tree. summary appears in the r console. notice summary information includes: # of tips and internal nodes, tip labels, and whether the tree is rooted or not
ggplot(tree) + geom_tree() #tree in ggplot with default theme. to clean up tree add the following theme
ggplot(tree) + geom_tree() + theme_tree() #ggplot with geom_tree() function and theme_tree
ggtree(tree) #short cut building a tree in ggplot with the tree theme. using this code will automatically build a geom_tree with theme_tree()

ggtree(tree) + theme_tree2() #this will build the above tree, but with the addition of a scale on the x axis. this scale can be used to represent the amount of genetic change)..length of branch represents evolutionary time

ggtree(tree, layout = "slanted") #different visualization

ggtree(tree, layout = "circular") #different visualization

ggtree(tree, branch.length = "none") #removing branch lenght makes a cladogram..not sure if we should include this
```

```{r}
?ggtree #we should also add "helpful R tips" throughout so they get used to finding answers to problems when they come up

#import your newick file into R and view a summary of your tree data to check that everything makes sense
tree <- read.tree("tree_newick.nwk") ##maybe we can use a differnet newick file so that they are exposed to two different layouts? just a thought..
tree 
#construct a tree using the ggtree + geom() or the shortcut you learned above
ggtree(tree) 

ggtree(tree) + theme_tree2() #this will build the above tree, but with the addition of a scale on the x axis. this scale can be used to represent the amount of genetic change)..length of branch represents evolutionary time




```


I think after we show them how to build a basic tree, we should walk through how to interpret different aspects of the tree. after we talk through branch lengths, nodes/ recent ancestors, phyletic groups, etc we could have some code for customizing their tree to focus on different clades or specific taxa.

