---
title: "206 tutorial- answers included"
authors: "Cheyenne M., Katie B., Emily"
date: "2/11/2019"
output: github_document
---

```{r setup, include=FALSE}
install.packages("knitr")
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

Before we begin, lets review the basic layout of RStudio. 
You are currently looking at the script window. You can enter text and save code chunks here.

The box below this one is called the console. The console is similar to the script window in that commands can be entered here but they will executed directly after you hit enter. 

The box in the upper right corner is the environment and history pane. The environment tab is where you will populate your data and values. The history tab is where you can track your previous work. 

The box on the bottom right has several tabs that contain different types of information. In this box is where you can find your files, install packages, and most importantly, where you can find help! 

You will find that all of these panels are very useful, however most of the work you do in this tutorial will be in this panel.

This tutorial is designed to help you construct a phylogenic tree and make some customizations that will make interpretation of the information it contains easier for you and your reader. You should know that RStudio and the tools you will be using are incredibly powerful and that this tutorial is in no way an exhaustive lists of their capabilities. That being said, once you complete this tutorial you should be well prepared to investigate these tools further! A great way to do this is to google your question (make sure you specify that you are working in ggtree) and the odds that someone has already answered this are high! There are plenty of forums where you can read about packages or functions, troubleshoot error messages, and ask for input on your code!

Now that you are familiar with the general layout of RStudio and how to find help when you need it, let's begin with a tutorial!


Guided Tutorial:
You may remember using Rstudio during Bio 205 to visualize and analyze data. Rstudio can also be used to visualize and customize phylogenic trees!

The following tutorial has been designed to walk you through the process of constructiong a phylogenic tree using R. You will learn about the different tools you will need to build a tree, how to customize a tree, and how to interpret information represented in a tree.

To begin, lets make sure that you have the tools you will need to make a tree!

R comes with many different types of tools that can be used to create and customize figures, as well as analyze your data and create meaningful images. Often these tools are created by users and organized into different "packages" for other users to explore and use in their own work. The first thing you will need to do is to "load" the tools you will need into Rstudio. This is done by using the command "install.packages()" as well as the command "library()" to load the tool you will use.
 
Install the tidyverse package and load the ggtree tool by clicking the green arrow in the upper right corner of the grey code chunk. This may take a few minutes to do. Grab a cup of tea and relax! Watch the console pane for updates on loading your packages. There will come a point when you are asked whether you want to update all/some/none of the packages. Type "a" in the console pane after this message and hit enter.
If a second Yes or No question appears, type Yes and hit enter.

```{r Loading packages}
install.packages("tidyverse")
install.packages("phytools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ggtree", version = "3.8")
```
```{r library packaages}
library(ggtree)
library(phytools)
```

As you click the green triangles to run code, you can see it running in the consile below. If your console is minimized, click "console" to expand it. This is also where it will ask you the questions mentioned above. 

The trees that you will be constructing were made using a trait table. A trait table is a way of tracking shared or disimilar traits between taxa and, when combined with your tree, to track when a certain trait arises. For example, from a the first trait table you will use, you can see which taxa have jaws, which taxa do not have feathers. We will get into this later in the tutorial, but for now lets get your data table formatted and into R.

To "load" your trait table into R, we will use the function "read.csv()". A csv file is a way of saving information that was formatted in an Excel spreadsheet to be read by R. The Excel spreadsheet has already been made for you and saved as a csv file. Now you need to tell R to "read" the file and asign it a name (let's call your trait table "data"). To do this, click the green arrow to in the upper right corner of the code chunk.

```{r trait table for test tree}
data <- read.csv("trait_data.csv", header= TRUE, sep= ",")
```
Take a look at your environment pane, do you see your new "data" under the data heading? Great! Click on your data in the environment window to view the trait table.


Now you will assign this newick file (a common text format for trees) the name text_string so you can use it to build a tree. The characters "<-" tells R that you want to assign information that you type to the right of it to the name that you type to the left of it. Now whenever you use the value "text_string", R will know that it includes the information in the newick file.

Assign the newick file to the name "text_string" now.

```{r tree text aka a newick file}
text_string <- "((((alligator, bald eagle),antelope), (seabass), lamprey));"
```
((((lamprey, sea bass),antelope), (bald eagle), alligator))
Now look in the environment panel. Notice that you have populated it with a new variable called "text_string" and it contains the newick file. Great job!


Now you need to show R how to interpret the information contained in your new value. To do this, you will use the command "read.tree()". This will make an "object" that now contains the information translated by R from the newick file into usable information for your tree! 

Use the command "read.tree()" to make an object called test_tree. This will be the object you use when plotting the tree. 

```{r reading tree}
test_tree <- read.tree(text=text_string)
```

Look back to the environment panel and click on the object "test_tree" to view the translated newick file.


Now that R is able to read the information you want to use, let's visualize this information into a tree! 

There are several functions for plotting phylogenic trees in R, but the simplest one is "plot()". This command works by typing the object that you want to plot followed by some formatting information."no.margin =" tells R to use the entire output frame to plot your tree and "edge.width =" tells R how thick to make the branches of your tree.

Plot your first tree now!
Running the code chunk below will create an image right below the chunk. 

```{r plotting tree}
plot(test_tree,no.margin=TRUE, edge.width=1)
```
Viola! you have a tree! 

This looks great, but let's make some alterations so you can explore the options that R gives you in tree making. 

First lets get a general idea of the information represented in your tree by having R generate a summary for you.

```{r summary of tree}
summary(test_tree)
```
Explore the informatin in the summary table. Notice that R can tell you how many tips and nodes are contained in your tree as well as what each tip label is. 


The "plot()" function is great for making simple trees, but if you want to get creative and customize your tree, it is much easier to use the function "ggtree()". This function allows you to highlight different parts of your tree, identify clades, find common ancestors as well as change the general aesthetics of your tree. First, lets use the same newick file that we translated above to make a basic ggtree. To add labels for our taxa we will use "+" and then a "geom_tiplab()" to label the tips.

Go ahead and make your tree using "ggtree()" and add labels to the tips of the tree usings "+ geom_tiplab()"
```{r Using ggtree to plot a tree}
ggtree(test_tree) + geom_tiplab() + xlim(0, 5)
```
Note: the xlim aregument helps to not cut off the tip labels. Change them around if you want to see what I mean, or get rid of the arguemnt entirely.

Great! Notice that this tree is essentially the same tree you made using the "plot()" function, but now you will see how easy it is to customize your ggtree!


First, lets start by identifying nodes within your tree. Identifying shared nodes will be useful later on when you want to find clades and other types of phyletic groups.
```{r Highlighting nodes}
ggtree(test_tree) + geom_tiplab(hjust = -.2) + geom_nodepoint(color="yellow", alpha=.9, size= 4) + xlim(0, 5)
```


Labelling a clade within a tree can be a really useful way to draw attention to a group of taxa that you want your reader to focus on. To highlight a clade we will use "+ geom_cladelabel()". Notice that there is some custom code within your clade label. "node =" tells R at what point in the tree to originate your clade. "label =" tells R what text you want to appear as your clade name and the following "color=" "offset=" and "align=" are different aesthetic parameters for your clade label. 

```{r How to label a clade}
ggtree(test_tree) + 
  geom_tiplab()+
  xlim(0, 6) +
  geom_cladelabel(node=8, label="Clade 1", 
                  color="red", offset=.8, align=TRUE) 
```
This looks great! Now your reader is able to interpret information about the tree you have made.


You may be wondering how to know what node number your clade originates from. To do this you will want to map your nodes. Think of this as a map to your tree, you probably would not use this tree in a publication or paper, but you will use it to help generate more descriptive trees. To do this, add ("+") node labels to your tree by using "geom_text()". The information contained within this function is telling R that you want to label the internal nodes of your tree ("aes(label=node) and to place the labels to the right of the nodes ("hjust= -.4").
```{r mapping nodes}
ggtree(test_tree) + 
  geom_text(aes(label=node), hjust=-.4)+
  xlim(0, 5)
```
Using your tree map, go back to the code chunk {r How to label a clade} and label the clade that only includes nodes 1 and 2 (sea bass and lamprey) by telling R which node to originate your clade from.


But what if you are not sure what the most recent commonly shared ancestor is for a group of taxa? Ggtree has some helpful functions for this! The function "mrca()" will generate a table where you can find the most recently shared node between any two taxa. For example, the most recently shared node between lamprey and sea bass is at node 9. 

You can also use the function "getMRCA()" to find the most recent common ancestor of two or more taxa. To use this function, simply tell R what tree it should look at ("phy= test_tree") and what taxa you want to know about (to do this use tip = c("taxa 1", "taxa 2",...)). 


```{r finding most recent common ancestors}
mrca(phy = test_tree, full = FALSE)

getMRCA(phy= test_tree, tip= c("lamprey", "seabass"))
```
In your output window you will see the MRCA table. Look at where lamprey and seabass intersect. Their most recently shared node is 9. You will also see that the output of the "getMRCA()" confirms this ([1] 9). 


You may also want to change the branch colors of a clade to make them stand out on your tree more. To do this, we have to tell ggtree the members of a clade using the function "groupClade(your tree, .node=# of your clades origin)" and assign that grouped clade to your tree using "<-". Then we will tell ggtree to make a new tree with all branches and members of that clade shown in a different color fromt the rest of your tree.
```{r grouping clades}
test_tree <- groupClade(test_tree, .node=9)
ggtree(test_tree, aes(color=group, linetype=group)) +
  geom_tiplab()+
  xlim(0, 5)

test_tree <- groupClade(test_tree, .node=c(9, 8))
ggtree(test_tree, aes(color=group, linetype=group)) + geom_tiplab(aes(subset=group==2)) + geom_tiplab()+
xlim(0, 5)
```
The code above produces 2 trees. To see them both use the arrows above them in the plots window. 

The code above is an easy and simple way of identifying and representing monoplytic groups on your tree! 


But what if you want to represent a paraphyletic or polypheletic group? What if you have a list of taxa, but you aren't sure if they are mono/para/polyphyletic? Ggtree has an answer for that as well! The function "groupOTU()" allows you to define wich taxa you want to investigate and will then trace back to their most recent common ancestor. Since you can define your taxa list however you like, the resulting group may be monophyletic (clade), polyphyletic or paraphyletic.

Let's try this out by investigating whether lamprey and antelopes belong to a monophyletic (clade), polyphyletic or paraphyletic group.

```{r identifying phylogenic groups}
test_tree <- groupOTU(test_tree, .node = c("lamprey", "antelope"))
ggtree(test_tree, aes(color=group, linetype=group)) +
  geom_tiplab()+
  xlim(0, 5)
```

What do you think? Are all offspring of the shared ancestor of antelope and lamprey represented in the blue grouping? What type of phyletic group does this tree show in blue then?


In addition to all of the wonderful ways that GGtree can customize your that you have just seen, it also allows you to make different types of tree! Though the focus of this tutorial is not to use these different layouts, it is nice to know that you could use them in the future. You could make a wonderful forests of unique phylogenic trees! The first layout, "slanted" is similar to the ggtree we have used above, but it smooths the branches to be aesthetically pleasing. The next two layouts are known as "unrooted" trees, which do not show direction of ancestoral relation like rooted trees do.
```{r Tree layouts}
ggtree(test_tree, layout="slanted") + geom_tiplab()+
  xlim(0, 5)
ggtree(test_tree, layout="circular")+ geom_tiplab()+
  xlim(0, 5)
ggtree(test_tree, layout="equal_angle")+ geom_tiplab()
```


Plant Assignment:

Now you will use the skills you learned from the guided tutorial to interpret and customize your own tree! An additional goal of this assignment is to help you understand your class material by working with data from your textbook and producing unique trees that highlight characteristics of your own choosing.

Here's the newick for the tree you will be using for the rest of this tutorial
Newicks are usually the output of tree building software like RAxML.

The tutorial will get you as far as building a basic tree, then you will need to use the code above to manipulate the tree!

```{r plant newick}
plants_newick <- "(Schmitzia_hiscockiana, (Nitellopsis_obtusa, ((Haplomitrium_cooperi, Blasia_pusilla, Porella_cordeana),((Sphagnum_cuspidata, Thuridium_delicatulum), ((Huperzia_lucidula, Lycopodium_digitatum), ((((Dryoteris marginalis, Dryopteris_intermedia),Dennstaedtia_punctilobula),((Equisetum_arvense)), ((Ginkgo_biloba, ((Pinus_strobilus, Pinus_rigida)), ((Baptisia_australis, Solanum_osscruentum,((Juglans_nigra, Carya_ovata)))))))))))));"
```
```{r reading plant newick}
plant_phylo <- read.tree(text=plants_newick)
```
```{r making basic plant phylo}
plot(plant_phylo,no.margin=TRUE, edge.width=1)
```

```{r making a basic plant phylo with ggtree}
plant_tree <- ggtree(plant_phylo) + geom_tiplab() + xlim(-10, 30)
plant_tree
```

So now you have basic tree with a bunch of scientific names. The goal is for you to use your skills from above, knowledge of trees, and likely google to figure out what exactly is on the tree and manipulate it to answer some questions. 


1. Where are the most primitive plants? The most evolved? Find them and highlight the branch of the most primitive group. 
```{r question 1}
plant_tree <- ggtree(plant_phylo) + geom_tiplab() + xlim(-10, 30)
#mapping nodes
ggtree(plant_phylo) +  
  geom_text(aes(label=node), hjust=-.4)

#highlighting most primitive 
plant_tree <- groupClade(plant_tree, .node=1)
ggtree(plant_tree, aes(color=group, linetype=group)) +
  geom_tiplab() + xlim(-10, 30)

#highlighting most evolved
plant_tree <- groupClade(plant_tree, .node=20)
ggtree(plant_tree, aes(color=group, linetype=group)) +
  geom_tiplab() + xlim(-10, 30)

#highlighting most primitive and most evolved on same tree
plant_tree <- groupClade(plant_tree, .node=c(1, 20))
ggtree(plant_tree, aes(color=group, linetype=group)) +
  geom_tiplab() + xlim(-10, 30)

```

2. Find the fern(s). What is the most closely related species. Put a cloud around ferns and its close relative(s). 
```{r question 2}
#to find mrca
getMRCA(phy= plant_phylo, tip = c(10, 11, 12))

#option 1
plant_tree <- groupClade(plant_tree, .node=c(10, 11, 12))
ggtree(plant_tree, aes(color=group, linetype=group)) + geom_tiplab(aes(subset=group==2)) + geom_tiplab() + xlim(-10, 30)

#option 2
plant_tree <- groupClade(plant_tree, .node=c(31))
ggtree(plant_tree, aes(color=group, linetype=group)) + geom_tiplab(aes(subset=group==2)) + geom_tiplab() + xlim(-10, 30)

```

3. Label the angiosperm clade.
```{r question 3}
ggtree(plant_tree)+
  geom_tiplab() +
  xlim(-7, 30) +
  geom_cladelabel(node=39, label= "Angiosperms", color="purple", offset= 12, align=TRUE)
```

4. Where did roots and/or tracheids evolve? Highlight the species they are present in.
```{r question 4}
plant_tree <- groupClade(plant_tree, .node=c(27))
ggtree(plant_tree, aes(color=group, linetype=group)) + geom_tiplab(aes(subset=group==2)) + geom_tiplab() + xlim(-10, 30)
```

5. Note all the places where vessel elements evolved and highlight them
```{r question 5}
ggtree(plant_tree)+ geom_tiplab() + xlim(-10, 30)+ 
  geom_nodepoint( color="brown", alpha=.9, size=3) + 
  geom_point2(aes(subset=(node == 28)), size=4, shape=23, fill="tan")+   geom_point2(aes(subset=(node == 31)), size=4, shape=23, fill="tan") + geom_point2(aes(subset=(node == 39)), size=4, shape=23, fill="tan")
```
ggtree(test_tree) + geom_tiplab(hjust = -.2) + geom_nodepoint(color="yellow", alpha=.9, size= 4) + xlim(0, 5)

6. Link the an angiosperm tree to a gymnosperm tree with green.
```{r question 6}
ggtree(plant_tree)+ geom_tiplab() + xlim(-10, 30)+ geom_taxalink('Carya_ovata', 'Pinus_strobilus', color='green')
```

7. Label the monophyletic groups at the base of the tree with the broad category they fall into. 
```{r question 7}
ggtree(plant_tree)+
  geom_tiplab() +
  xlim(-7, 30) +
  geom_cladelabel(node=1, label= "red algae", color="red", offset= 12, align=TRUE)+
  geom_cladelabel(node=2, label="green algae", color="green", offset=12, align=TRUE)
```

Here is a tree of common names too, in case you think it should be provided
```{r common name newick}
common_name_newick <- "(red_algae, (green_algae, ((liverwort1, liverwort2, liverwort3),((Sphagnum_moss, fern_moss), ((shining_clubmoss, running_cedar), ((((marginal_woodfern, intermediate_woodfern),hay-scented_fern),((horsetail)), ((Ginkgo, ((white_pine, pitch_pine)), ((false_indigo, australian_eggplant,((black_walnut, shagbark_hickory)))))))))))));"
```
```{r reading cn newick}
cn_phylo <- read.tree(text=common_name_newick)
```
```{r making cn tree}
plot(cn_phylo,no.margin=TRUE, edge.width=1)
```


If you get stuck, google is your friend! Rememeber to include terms like ggtree. Also, this FAQ website for the package is very useful:
https://guangchuangyu.github.io/software/ggtree/faq/

Other resources used: 
https://4va.github.io/biodatasci/r-ggtree.html
http://bioconductor.org/packages/release/bioc/vignettes/treeio/inst/doc/Importer.html
https://bioconductor.org/packages/release/bioc/vignettes/ggtree/inst/doc/treeManipulation.html
https://www.khanacademy.org/science/biology/her/tree-of-life/a/building-an-evolutionary-tree
BIOL 205 R tutorial 
Animal Behavior and Neuro tutorial
Dr. Field