---
title: "Final 206"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the document where we will put all of the code we know we want to use for the BIOL 206 tutorial, as well as detailed descriptions of what the code is doing etc. 

The first section id to get you aclimted to using R to look at trees and manipulate them. 

Test Tree

Reading in a table from a .csv (a type of excel file)
```{r character table for test tree}
data <- read.csv("trait_data.csv", header= TRUE, sep= ",")
```

Now you will assign this newick file (a common text format for trees) the name text_string so you can use it to build a tree 
```{r tree text aka a newick file}
text_string <- "((((lamprey, sea bass),antelope), (bald eagle), alligator));"
```

Use the comman read.tree to make an object called test_tree. This will be the object you use when plotting the tree. 
```{r reading tree}
test_tree<- read.tree(text=text_string)
```

This is the simplest tree plotting function. 
```{r plotting tree}
plot(test_tree,no.margin=TRUE, edge.width=2)
```
Viola! you have a tree! 

Let's make some alterations so you can explore the options that R gives you in tree making. 

summary 
label nodes
highlight branches
cloud groups








